Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-08 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: tried to improve stacking discussion

% 1. Description of data and assumption o \ell being shared
Suppose we have a set of $N$ GW events $\{d_{i}\}$ that are described by
model parameters $\bm{\lambda}_{i}$, where $i$ labels each event.
%
We assume that the value of the coupling constant $\ell$, given a specific theory,
is common to all events.
%
% 2. Define joint posterior
%
The joint posterior for the parameters $\bm{\lambda}_{i}$, that describe the
data $\{d_{i}\}$, is
%
\begin{equation}
P(\ell, \{\btheta_i\} | \{d_{i}\})
= \frac{ P(\ell_i, \{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ) }{ P(\{d_i\}) }.
\label{eq:def_joint_posterior}
\end{equation}
%
where we separated $\ell$ from the GR waveform model parameters $\btheta$ [cf.~Eq.~\eqref{eq:def_params}].
%
% 3. Marginalisation to obtain the cumulative posterior
%
To obtain the cumulative posterior distribution on $\ell$, we marginalize
the joint posterior~\eqref{eq:def_joint_posterior} over
all parameters $\{ \btheta_i \}$, i.e.,
%
\begin{align}
P(\ell | \{d_i\}) &= \int \dd \{\btheta_i\} \, P(\ell, \{\btheta_i\} | \{d_{i}\}),
\nonumber \\
                  &= \int \dd \{\btheta_i\} \, \frac{ P(\ell_i, \{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ) }{ P(\{d_i\}) },
\nonumber \\
                  &= \frac{P(\ell_i)}{P(\{d_i\})} \int \dd \{\btheta_i\} \, P(\{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ),
\end{align}
%
where we have assumed that the prior on $\ell$, $P(\ell)$, is independent of
the other parameters $\{\btheta_i\}$.
%
This allow us to factor $P(\lambda, \{\btheta_i\} = P(\ell) P(\{\btheta\})$
and then move $P(\ell)$ out of the integral
%
We also moved out integral the evidence $P(\{d_i\})$, which is a normalization constant.
%
% 4. Factor the integrals in a product due to statistical independence
We can now factor the probabilities inside the integral, because each event is independent from one another:
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell_i)}{P(\{d_i\})}
\prod_{i=1}^{N}
\int \dd \btheta_i \, P(\btheta_i) \, P( d_i | \ell_i, \btheta_i ).
\end{equation}
%
% 5. Final formula
We can now identify the integral as the marginalized likelihood for each individual event
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell_i)}{P(\{d_i\})}
\prod_{i=1}^{N} P( d_i | \ell_i ),
\end{equation}
%
which we can rewritte as
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell_i)}{P(\{d_i\})}
\prod_{i=1}^{N} \frac{P(d_i)}{P(\ell_i)} P(\ell_i | d_i).
\label{eq:cumulative_dist_ell}
\end{equation}
%
using Bayes' theorem. This is our final result.

In our parameter estimation studies, we found that depending on which event and
on which theory we adopt for our parameter estimation, that we had to adjust
the prior on $\ell$.
%
The reasons are twofold. First, we found that depending on the source parameters $\btheta$,
certain large values of $\ell$ could result in unphysical waveforms in the frequency band we considered.
For instance, they could feature a ringdown multiple times longer than the inspiral-merger part.
%
Such waveform are perhaps not surprising since we are, by construction, adding corrections to the GR QNM
which \emph{delay} its damping.
%
%
In practice, we found these edge cases were problematic to deal with \texttt{LALInferenceMCMC}.
%
We addressed this issue, by reducing the prior range on $\ell$, which does not pose a problem to our parameter
estimation runs because these edge-case waveforms differ grossly from the true signal and thus have
negligible likelihood
%
However, we cannot reduce the prior on $\ell$ too much. If we do so,
to the individual posterior to rail against the prior and to result in GW
waveform which are physical.
%
Altogether, this means that the priors $P(\ell)$ are not shared between events and
they must be taken into account when calculation the cumulative posterior function~\eqref{eq:cumulative_dist_ell}.

To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is relies on first obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. Next, we multiply
these posteriors together to obtain the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.

>>>>

% 1. Description of data and assumption o \ell being shared
Suppose we have a data set $\{d_{i}\}$ of $N$ events which is described by
model parameters $\bm{\lambda}_{i}$, where $i$ labels each event.
%
We assume that the value of the coupling constant $\ell$, given a specific theory,
is common to all events.
%
% 2. Define joint posterior
%
The joint posterior for the parameters $\bm{\lambda}_{i}$, that describe the
data $\{d_{i}\}$, is
%
\begin{equation}
P(\ell, \{\btheta_i\} | \{d_{i}\})
= \frac{ P(\ell, \{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ) }{ P(\{d_i\}) }.
\label{eq:def_joint_posterior}
\end{equation}
%
where we separated $\ell$ from the GR waveform model parameters $\btheta$ [cf.~Eq.~\eqref{eq:def_params}].
%
% 3. Marginalisation to obtain the cumulative posterior
%
To obtain the cumulative posterior distribution on $\ell$, we marginalize
the joint posterior~\eqref{eq:def_joint_posterior} over
all parameters $\{ \btheta_i \}$, i.e.,
%
\begin{align}
P(\ell | \{d_i\}) &= \int \dd \{\btheta_i\} \, P(\ell, \{\btheta_i\} | \{d_{i}\}),
\nonumber \\
                  &= \int \dd \{\btheta_i\} \, \frac{ P(\ell, \{\btheta_i\}) \, P( \{d_i\} | \ell_i, \{\btheta_i\} ) }{ P(\{d_i\}) },
\nonumber \\
                  &= \frac{P(\ell)}{P(\{d_i\})} \int \dd \{\btheta_i\} \, P(\{\btheta_i\}) \, P( \{d_i\} | \ell, \{\btheta_i\} ),
\end{align}
%
where we have assumed that the prior on $\ell$, $P(\ell)$, is independent of
the other parameters $\{\btheta_i\}$.
%
This allow us to write $P(\ell, \{\btheta_i\} = P(\ell) P(\{\btheta_i\})$
and then move $P(\ell)$ out of the integral
%
We also moved out integral the evidence $P(\{d_i\})$, which is a normalization constant.
%
% 4. Factor the integrals in a product due to statistical independence
%
We can now factor the probabilities inside the integral, since each event is independent from another:
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell)}{P(\{d_i\})}
\prod_{i=1}^{N}
\int \dd \btheta_i \, P(\btheta_i) \, P( d_i | \ell_i, \btheta_i ).
\end{equation}
%
% 5. Final formula
We can now identify the integral as the marginalized likelihood for each individual event
%
\begin{equation}
P(\ell | \{d_i\}) = \frac{P(\ell)}{P(\{d_i\})}
\prod_{i=1}^{N} P( d_i | \ell_i ),
\end{equation}
%
which we can rewrite as
%
\begin{align}
P(\ell | \{d_i\}) &= \frac{P(\ell)}{P(\{d_i\})}
\prod_{i=1}^{N} \frac{P(d_i)}{P(\ell_i)} P(\ell_i | d_i)
\nonumber \\
&= P(\ell)
\prod_{i=1}^{N} \frac{P(\ell_i | d_i)}{P(\ell_i)}.
\label{eq:cumulative_dist_ell}
\end{align}
%
by using Bayes' theorem in the first line and cancelling the evidences in the
second line. This is our final result.

%
\hs{Abhirup, maybe move this to the prior subsection?}
In our parameter estimation runs that we had to adjust our prior on $\ell$ depending on which event and
on which theory we examined.
%
The reasons are twofold.
%
% Describe problem 1
%
First, we found that depending on the source parameters $\btheta$,
large values of $\ell$ could result in unphysical waveforms in the frequency
band we considered. More specifically, these waveforms have a ringdown which is
considerable longer than the inspiral-merger part in the detector frequency band.
%
The fact that our sampled over these waveforms is perhaps not surprising,
because, after all, we are by construction adding corrections to the GR QNM
which \emph{increase} the damping timescale.
%
In practice, we found that these edge-cases were problematic to deal with
\texttt{LALInferenceMCMC}.
%
We overcome this problem, we reduced the prior upper-value of our prior on $\ell$.
%
This choice does not pose a problem to our parameter estimations, because these
waveforms differ grossly from the data and thus have negligible likelihood.
%
% Describe problem 2
%
However, we cannot reduce the prior on $\ell$ too much. If we do so,
to the marginalized posterior distribution on $\ell$ would rail against
the prior or, in a more extreme case, even be a uninformative, flat posterior.
%
% Summary
%
Altogether, this means that the priors $P(\ell)$ are not shared between events and
they must be taken into account when calculation the cumulative posterior function~\eqref{eq:cumulative_dist_ell}.

Finally, we use one of the approaches of Ref.~\cite{Perkins:2021mhb} to
compute the product in Eq.~\eqref{eq:cumulative_dist_ell}.
%
It method relies on first obtaining a kernel density estimator (KDE) for
the posterior distribution on $\ell$ for each individual event.
%
Next, we multiply these posteriors according to Eq.~\eqref{eq:cumulative_dist_ell}.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This approach based on the KDE can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0, \infty]$.
